\paperTitle

### 2014-01-31 doc/icml2014/undirected.tex
done with undirected

\section{Consistent Parameter Estimation for Undirected Latent Variable Log-linear Models}
\label{sec:undirected}

In this section, we will show that we can extend the results to
  undirected graphical models parameterized as log-linear latent variable
  models.
We exploit a connection between the moments of the clique potentials and
  the log-partition function to show that learning parameters is convex
  given the clique marginals $Z_\sC$. 
The end result is a consistent estimator for log-linear
  latent-variable models which satisfy \propertyref{exclusive-views}.
Finally, we generalize the result to classes when only a subset of the
  clique marginals can be learned.

As before, let $\sG$ be a graphical model with variables $V$, with $X
  \subseteq V$ being observed and $H = V \setminus X$ hidden.
We assume that the undirected graphical model is a log-linear latent
  variable model with parameters $\theta$,
\begin{align*}
  p_\theta(\vx, \vh) &= \exp( \theta^T\phi(x,z) - A(\theta) ),
\end{align*}
where $\phi(x,z)$ is a feature vector with components $\phi_\sC(x,z)$
  for every clique and $A(\theta) = \sum_{\vx, \vh}  \exp(
  \theta^T\phi(x,z) )$ is the log-partition function.
While $\phi_\sC(x,z)$ can be an arbitrary featurization, we assume
  without loss of generality that $\phi_\sC(x,z)$ are the indicator
  features $\BI[x,z]$\footnote{Note that $\BI[x,z]$ are sufficient
  statistics for the model, and hence we do not lose any statistical
  efficiency in using them}.

As before, our objective is to find a set of parameters $\hat \theta$
  that maximize the likelihood of data $\sD$.
\begin{align}
  \hat \theta 
      &= \arg\max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \nonumber \\
      &= \arg\max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh} p_\theta(\vx,\vh) \nonumber \\
      &= \arg\max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh} \exp( \theta^T\phi(\vx,\vh) - A(\theta) ). \label{eqn:obj-ml}
\end{align}
Note that this problem is not concave in $\theta$.

An alternate objective is to optimize the following lower-bound on the log-likelihood, 
\begin{align}
  \sL_p(q,\theta) &= \sum_{\vx \in \sD} \sum_{\vh} q(\vh \given \vx) \log\frac{p_\theta(\vx,\vh)}{q(\vh \given \vx)} \\
    &= \KL( q(\vh \given \vx) || p_{\theta}(\vh \given \vx)) \label{eqn:obj-kl}
\end{align}

Importantly, given $q$, the above objective is convex in $\theta$ and vice versa.
It can be shown\cite{} that the local-optima of \equationref{obj-kl}
are also local-optima of \equationref{obj-ml}. Usually, one
iteratively optimizes $q$ and $p_\theta$ from some initial parameters
$\theta_0$ to arrive at a local optimum.

This reduces to,
\begin{align}
  \sL_p(\theta) &= \theta^T Z - A(\theta),
\end{align}
which is a convex problem.

However, from the previous section, for any model satisfying
  \property{exclusive-views}, we have a consistent estimator for the
  marginals $q \equiv Z$. 
Therefore, we can optimize for $\theta$ given this estimate of $q$,
  $\hat q$.

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnLogLinear}
  \label{algo:undirected}
  \begin{algorithmic}
    \REQUIRE A graphical model $\sG$ satisfying \textbf{(P1)}, \textbf{(P2)}, data $\sD$
    \ENSURE A consistent estimate $\hat \theta$ of the parameters of the model.
      \STATE Apply \LearnMarginals to $\sG$ to learn clique marginals $Z_\sC$.
      \STATE Optimize \equationref{obj-kl} with respect to $\theta$.
  \end{algorithmic}
\end{algorithm}

This approach is outlined in \algorithmref{undirected}.

\todo{Do we have to worry about the fact that this is constrained?}
\todo{What about model un-identifiability?}

\paragraph{Learning with partial moment constraints}

In more general models, \propertyref{exclusive-views} may not hold for
  every clique in the graph, however we can still estimate the marginals
  $Z_\sC$ for cliques in which it does.
In this section, we will describe a natural method to use these
  constraints to simplify the parameter search problem.

Constrain the KL to be in $Q$.

\subsection{Examples}

\paragraph{Latent MRFs}

Image segmentation

Separate features into views

\todo{Some challenge example requiring measurements em}

>>>>

\section{Undirected log-linear models}
\label{sec:undirected}

% DONE: need more of a transition from previous sections
Recall that we used tensor factorization and composite likelihood
to estimate the latent marginals $Z_\sC = \BP(\vh_\sC)$ and $\mOpp{v}{i} = \BP(x_v \mid h_i)$.
For directed models, obtaining the local conditional tables from $Z_\sC$ requires
renormalizing $Z_\sC$.
For undirected log-linear models, the parameters cannot simply be obtained locally;
we will have to perform global optimization.
Fortunately, as we will show, this optimization problem is convex,
by relating the recovered conditional moments with the sufficient statistics of
the log-linear model.

% Define model
As before, let $\sG$ be a graphical model with observed variables $X$ and hidden variables $H$
(see \figureref{examples-mrf} for an example).
For each clique $\sC \in \sG$, we have a feature vector $\phi_\sC(\vx_\sC, \vh_\sC)$,
where $\vx_\sC$ and $\vh_\sC$ are the observed and hidden variables in $\sC$, respectively.
Let $\phi(\vx, \vh) = \sum_{\sC \in \sG} \theta^\top \phi_\sC(\vx_\sC,\vh_\sC)$ be the global feature vector.
We consider log-linear models of the following form:
\begin{align}
p_\theta(\vx, \vh) &= \exp\left( \theta^\top \phi(\vx,\vh) - A(\theta) \right),
\end{align}
where $A(\theta) = \sum_{\vx, \vh}  \exp( \sum_\sC \theta^\top \phi(\vx,\vh) )$ is the log-partition function.

% FIGURE: example
\begin{figure}
  \label{fig:examples-mrf}
  \centering
  \subimport{figures/}{mrf.tikz}
  \caption{Example: undirected grid model where each hidden variable has two
  conditionally independent observations.
  We derive a consistent estimate.}
\end{figure}

%As an example, consider an undirected grid model in \figureref{examples-mrf},
%where each hidden variable has two conditionally independent observations.
%In practice, if two features are known to be independent, then they can
  %be separated into independent observed variables.
%Let the parameters of the model be $O \eqdef \Phi(x^a_{i,j}, h_{i,j})
  %= \Phi(x^b_{i,j}, h_{i,j})$ and $T \eqdef \Phi(h_{i,j}, h_{i \pm 1,j \pm 1})$. 
%Note that as this is an undirected model, the clique marginals do not
  %correspond to the parameters of the model.
%To recover the parameters of this model, note that every hidden
%  variable, $h_{ij} \in \sH$ is a bottleneck, with views
%  $x^{a}_{ij},x^{b}_{ij}$ and any other observed variable in the graph. 
%Thus, \LearnMarginals will be able to learn the marginals of each clique
%  and \LearnLogLinear will give us a consistent estimate of the
%  parameters.

Of course, the log-likelihood is non-concave:
\begin{align}
L_\text{unsup}(\theta) \eqdef \E_{\vx \sim \sD}[\log \sum_{\vh \in \sH} p_\theta(\vx,\vh)].
\end{align}
% PL: this doesn't make sense
%While $\phi_\sC(x,h)$ can be an arbitrary featurization, we assume
%  without loss of generality that $\phi_\sC(x,h)$ are the indicator
%  features $\BI[x,h]$\footnote{Note that $\BI[x,h]$ are sufficient
%  statistics for the model, and hence we do not lose any statistical
%  efficiency in using them.}.

%In this section, we will show that we can extend the results to
  %undirected graphical models parameterized as log-linear latent variable
  %models.
On the other hand, 
if we were able to observe $\vh$, the \emph{supervised} likelihood would be concave:
\begin{align}
\label{eqn:logLinearSupervised}
L_\text{sup}(\theta) &\eqdef \E_{(\vx,\vh) \sim \sD_\text{sup}}[\log p_\theta(\vx,\vh)] \\
                     &= \theta^\top \left(\sum_{\sC \in \sG} \E[\phi(\vx_\sC,\vh_\sC)]\right) - A(\theta).
\end{align}
While we don't have supervised data,
note that the method of moments gives us latent marginals $Z_\sC = \BP(\vh_\sC)$.
Furthermore, given $\mOpp{v}{i}$ and $Z_\sC$ of any $\sC \ni v$, we can recover
all marginals $Z_\sC = \BP(\vx_\sC, \vh_\sC)$ (with a slight augmentation of notation).
Finally, the key step is that we can compute all the expected feature vectors from the marginals:
\begin{align}
\label{eqn:logLinearFeatures}
\mu_\sC \eqdef \E[\phi(\vx_\sC,\vh_\sC)] = \sum_{\vx_\sC,\vh_\sC} Z_\sC(\vx_\sC,\vh_\sC) \phi(\vx_\sC,\vh_\sC).
\end{align}
Therefore, we can optimize the supervised likelihood objective without actually
having supervised data.
Of course, in the finite data regime, the method of moments yields estimates
$\hat \mu^\text{mom}_\sC$ which approximate the true $\mu_\sC$.
In supervised learning, we have different estimate $\hat\mu^\text{sup}_\sC$ of $\mu_\sC$ based on an empirical average
over the data points.
In the limit of infinite data, both estimators converge to $\mu_\sC$.

%We exploit the connection between the moments of the clique potentials and
  %the log-partition function to show that learning parameters is convex
  %given the clique marginals $Z_\sC$. 
%The end result is a consistent estimator for log-linear
  %latent-variable models which satisfy \propertyref{exclusive-views}.
%Finally, we generalize the result to classes when only a subset of the
  %clique marginals can be learned.

%As before, our objective is to find a set of parameters $\hat \theta$
%  that maximize the likelihood of data $\sD$.
%\begin{align}
%  \hat \theta 
%      &= \arg\max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \nonumber \\
%      &= \arg\max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh) \nonumber \\
%      &= \arg\max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} \exp( \theta^\top\phi(\vx,\vh) - A(\theta) ). \label{eqn:obj-ml}
%\end{align}
%Note that this problem is not concave in $\theta$. 
%
%\note{Consider the supervised setting.
%As we get infinite amounts of data, we get this simple form.
%As MoM is a consistent estimator for marginals, we can optimize this
%  likelihood and it is consistent.
%}

%However, given the moments $\E[ \phi_\sC(\vx,\vh) ] = Z_\sC$, the
%  problem simplifies to,
%\begin{align}
%  \hat \theta 
%  &= \arg\max_{\theta} \sum_{\sC \in \sG} \theta_\sC^\top Z_\sC - A(\theta), \label{eqn:obj}
%\end{align}
%which is concave in $\theta$.
\algorithmref{undirected} summarizes our approach.
In \sectionref{directed}, we showed that $\LearnMarginals$ is a consistent
estimator for $Z_\sC$ for any graphical model satisfying the exclusive views property (\propertyref{exclusive-views}). 
Consequently, the result is that the output of \algorithmref{undirected},
$\hat\theta$, is a consistent estimator.\footnote{Of course, the log-linear model must be identifiable
(violated, for example, if two components of $\phi(\vx_\sC, \vh_\sC)$ were identical).
In general, we can identify the parameters up to kernel of $\nabla^2 A(\theta)$.}

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnLogLinear}
  \label{algo:undirected}
  \begin{algorithmic}
    \REQUIRE A graphical model $\sG$ satisfying \propertyref{bottleneck}, data $\sD$
    \ENSURE A consistent estimate of parameters $\hat \theta$
      \STATE Apply $\LearnMarginals$ to $\sG$ to learn clique marginals $Z_\sC$.
      \STATE Compute expected features (\equationref{logLinearFeatures}) and optimize \equationref{logLinearSupervised}.
  \end{algorithmic}
\end{algorithm}

\paragraph{Pseudolikelihood.}
Unlike for directed models, in the undirected case,
graphs with high tree-width present a challenge
for even fully-supervised maximum likelihood.
This motivates looking at pseudolikelihood objective of \citep{besag75pseudo},
which optimizes the likelihood of each variable $a$ given its neighbors $\sN(a)$:
$\sum_{a \in X \cup H} \log p_\theta(a \mid x_{\sN(a)})$.
In the fully-supervised setting, pseudolikelihood provides consistent estimates,
albeit at the cost of statistical efficiency.
However, if we write the objective, we will see that the log-partition function $A(\theta)$
is replaced with a sum over $A_a(\theta; x_{\sN(a)}) = \sum_a \exp(\theta^\top\phi(a))$,
which cannot be computed from $Z_\sC$, since the neighbors $\sN(a)$ in general do not form a clique.
However, we adapt our earlier algorithm to estimate the marginals over all neighborhoods $Z_{\sN(a)}$.
Of course, this method is only viable for graphs of low degree ($|\sN(a)|$ is small).
For example, in the grid model, the degree is only 4.
Note that neither sample nor computational complexity depends on the treewidth.

\paragraph{Learning with partial moment constraints.}

In general, we might only have exclusive views (\propertyref{exclusive-views}) for a subset of the cliques.
In this case, we can estimate the expected features $\mu_\sC$ for those cliques
and use posterior regularization \citep{graca08em} or generalized expectation criteria \citep{mann08ge}
to encourage $\E_{p_\theta}[\phi(\vx_\sC,\vh_\sC)]$ to match $\mu_\sC$.
The resulting objective functions are still non-convex, but in practice we observed
that the extra information $\mu_\sC$ helps alleviate local optima.
%We can then integrate these constraints using posterior regularization \citep{graca08em}:
%\begin{align}
%L_\text{pr}(\theta) \eqdef L_\text{unsup}(\theta) + C\min_{q \in \sQ} \text{KL}(q(\vx,\vh) \| p_\theta(\vx,\vh)).
%\end{align}
%The main idea is to regularize the parameters towards values that produce the 
%While the optimization problem \eqref{eqn:obj-ml} is non-convex, we
  %expect that the measurements aid recover of parameters, and indeed see
  %this is true for experiments on simulated data.
%\todo{Include figure}

%Posterior regularization optimizes the following objective,
%\begin{align*}
%  \hat \theta &= \arg\min_{\theta} \arg\min_{q \in \sQ} \KL( q(\vx,\vh) \| p_{\theta}(\vx, \vh) ),
%\end{align*}
%where $q$ is a marginal distribution that belongs in constrained set
%$\sQ$. In our case, $\sQ$ is constrained to agree with the partial
%moments we recovered using \LearnMarginals, i.e. 
%\begin{align*}
%  \sQ &= \{ q(\vx,\vh) \given \forall \sC \in \sP ~ \E_{q(\vx,\vh)}[\phi_\sC(\vx, \vh)] = Z_\sC \}.
%\end{align*}

%In practice, we would like to regularize for error in our
%  estimates of $Z_\sC$; to do so, we use the measurements model of
%  \citet{liang09measurements} which defines a Bayesian model for the noise
%  in the marginals $Z_\sC$. The resultant optimization problem is,
%\begin{align*}
%  \hat\theta 
%  &= \arg\min_\theta - \sum_{x \in \sD} \log p_\theta(x) + \half \eta_\theta \| \theta \|^2 \\
%  &\quad + \half \eta_\beta \| Z_\sC - \sum_{\vx \in \sD} \E_{\theta}[\phi_\sC(\vx,\vh)] \|^2,
%\end{align*}
%where $\eta_\theta$ and $\eta_\beta$ are regularization parameters.

%
%However, the objective \equationref{obj} requires us to compute the
%  partition function, $A(\theta)$ and its gradient. 
%This scales with the treewidth of the model, which is the number
%  of rows/columns of the grid in this example, and hence can be
%  intractable.
%A more computationally efficient, but still consistent, alternative is
%to optimize the pseudo-likelihood \todo{how do I write this out}.
